Spending too much time at the Galactic bar: Chaotic evolution of the Ophiuchus stream

### 2015-11-23 ms.tex
remove old title

%\title{Tapping the keg: the Galactic bar peculiarity of the Ophiuchus stream and the Galactic bar}
\title{Spending too much time at the Galactic bar: Chaotic evolution of the Ophiuchus stream}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
            Branimir Sesar\altaffilmark{\mpia}, 
            Kathryn V. Johnston\altaffilmark{\colum},
            Hans-Walter Rix\altaffilmark{\mpia}
}

@@

We have found that a qualitative but observationally-motivated potential model for the Galactic bar generates a significant number of chaotic orbits in the vicinity of the recently-discovered Ophiuchus stream. By sampling many possible phase-space 6D positions and velocities for the stream centroid from the observational error distribution, we have shown that most possible orbits are strongly chaotic in the time-dependent, barred potential model for the Milky Way, but are mostly regular, eccentric orbits in a static, axisymmetric potential. We have shown that---using \streakline\ models of the stream density---the apparent shortness of the stream and the existence of nearby, high-dispersion stars can be explained by chaotic density evolution of the stripped stars. This is the first time chaos has been used to explain the morphology and [...] of a stellar stream. This work demonstrates the importance of the Galactic bar on the potential of the inner Milky Way and has important implications for future modeling of streams near in this region. This result also motivates deep imaging around the Ophiuchus stream to measure the density contrast of the ``fanned'' debris compared to the main stream and to find more candidate stars associated with chaotic tidal debris.

>>>>

\title{Spending too much time at the Galactic bar: Chaotic evolution of the Ophiuchus stream}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
            Branimir Sesar\altaffilmark{\mpia}, 
            Kathryn V. Johnston\altaffilmark{\colum},
            Hans-Walter Rix\altaffilmark{\mpia}
}

@@

We have found that, with a qualitative but observationally-motivated potential model for the Galactic bar, the orbits of the Ophiuchus stream stars are likely sensitive to the time-dependence and shape of the bar potential. For modeling the stream itself, it is therefore crucial to include this component of the Galactic potential. By fitting orbits to kinematic data for members of the stream in Milky Way-like potential models, we have found that orbits in the vicinity of the Ophiuchus stream are strongly chaotic for a range of bar parameters (pattern speeds and present-day angles). Using mock stellar stream models generated assuming a globular cluster-mass progenitor object, we have shown that the apparent shortness of the stream and the existence of nearby stars with very high velocity dispersion are plausibly explained by chaotic density evolution of the stars stripped from the progenitor object.

This is the first time chaos has been used to explain the morphology of a stellar stream and the first observational evidence for the importance of chaos in the Galactic halo. It also highlights the importance of including the Galactic bar in dynamical modeling of the Milky Way's inner halo and has important implications for future modeling of streams near in this region. With more Ophiuchus stream members, density and velocity information over a larger region near the stream, and better models for the internal structure of the Galactic bar, careful modeling of this stream could lead to tight constraints on the time-dependent properties of the bar.
